### 2020-09-30 hpp.tex
Write related work.

\section{Introduction}
This \paper describes a software platform taylored for manipulation planning.
The main contributions are:
\begin{itemize}
\item an original and general implementation of prehensile manipulation based on
  non-linear constraints,
\item an original solver for non-linear constraints that is able to handle implicit and explicit constraints,
\item a manipulation planning algorithm that is able to tackle a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above following state of the art development tools and methods.
\end{itemize}

The \paper is organized as follows.
\partie~\ref{partie:preliminary} introduces some preliminary notions like kinematic chains and Lie groups that are used to model the configuration space of each joint.

Each \partie is implemented by one or several software packages. Instead of providing formulas for some values need to be computed, we provide a link to the C++ implementation.

@@

\paragraph{Vector representations}
Each Liegroup element is represented by a vector. Rotations are represented by
unit quaternions.

@@

\section{Non-linear constraints and solvers}

@@

To optimize constraint resolution, we perform variable substitution when possible in order to reduce both the number of variables and the dimension of the resulting implicit constraint. We describe here the method that has been first published in~\cite{mirabel:hal-01804774}. Unlike in the former paper, the description we give in Algorithm~\ref{alg:ExplicitConstraintSet} is closer to the real implementation. Some links to the source code are indeed provided in the algorithm description.
\begin{algorithm}
  \begin{algorithmic}[1]
    \Procedure{initializeSolver}{}
    \State{$explicit$} $\gets$ empty vector of explicit constraints
    \State{$nc \gets 0$}
    \State{$args \gets$ array of size $nq$ filled with -1}
    \EndProcedure
    \Function{add}{$E=(in, out, f)$}
    \If{$args_{out}$ contains an element $\geq 0$}
    \State{return failure}
    \EndIf
    \State{queue $idxArg \gets$ elements of $in$}
    \While{$idxArg$ \textbf{not} empty}
    \State{$iArg \gets idxArg$ first element}
    \State{remove $idxArgs$ first element}
    \If{$iArg\in out$}
    \State{return failure}
    \EndIf
    \If{$args[iArg] == -1$}
    \State{continue}
    \Else
    \State{push $explicit[args[iArg]].in$ elements into $idxArg$}
    \EndIf
    \EndWhile
    \State{fill $args_{out}$ with $nc$}
    \State{explicit.add($E$); $nc\gets nc+1$}
    \State{computeOrder()}
    \State{\Return{success}}
    \EndFunction
  \end{algorithmic}
  \caption{Insertion of an explicit constraint in the solver. Line 1 is called once at initialization of the solver. $explicit$ is a vector that stores the constraints that are successfully added to the solver. $nc$ is the size of the latter. $args$ is an array that stores for each configuration variable, the index in $explicit$ of the constraint that computes this configuration variable, -1 if no constraints computes the index. Procedure \href{https://github.com/humanoid-path-planner/hpp-constraints/blob/2555dfec945575f824bd973c30c5c13ec3b67645/src/explicit-constraint-set.cc\#L144}{ADD} tests whether explicit constraint $E$ is compatible with the previously inserted constraints. Line 6 checks whether any output variable of $E$ is already computed by a previous explicit constraint. If so the procedure returns failure and the $E$ is not inserted. The Loop at line 9 recursively checks that any element of $out$ is not an input variable of a previously inserted constraint. If the loop ends without returning failure,  line 18 stores that elements of $out$ are computed by $E$ and $E$ is inserted in vector of constraints. Function \href{https://github.com/humanoid-path-planner/hpp-constraints/blob/2555dfec945575f824bd973c30c5c13ec3b67645/src/explicit-constraint-set.cc\#L301}{computeOrder} at line 20 recursively computes the order in which the explicit constraints are evaluated, following the rule that the input of a constraint should be evaluated before the output.}
  \label{alg:ExplicitConstraintSet}
\end{algorithm}
Once several compatible explicit constraints have been inserted in the solver,
they behave as a single one. For instance, if $\conf=(\conf_1,\conf_2,\conf_3)$,
$$
  \left\{\begin{array}{lcl}\conf_1 &=& f_1(\conf_2)\\
  \conf_2 &=& f_2(\conf_3)\end{array}\right. \mbox{ becomes }
  \left[\begin{array}{cc}\conf_1\\ \conf_2\end{array}\right] =
    \left[\begin{array}{cc}f_1(f_2(\conf_3))\\ f_2(\conf_3)\end{array}\right],
$$
and $f_2$ should be evaluated before $f_1$.

@@

\section{Manipulation Problem}

@@

We denote by $o_i(\conf)$ and $f_j(\conf)$ the poses in configuration $\conf$ of frames with respective centers $C_{o_i}$ and $C_{f_j}$ and with $x$-axis normal to each polygon. In an analogous way as in Definition~\ref{def:grasp}, we define
\begin{equation}\label{eq:convex-shape-contact-hbar}
\bar{h}(\conf) = \log_{\reals^3\times SO(3)} \left(f_{j}(\conf)^{-1}o_i(\conf)\right)
\end{equation}
The contact constraint is defined by the following piecewise differentiable function:
\begin{equation}\label{eq:convex-shape-contact}
h(\conf) = \left\{\begin{array}{lll}(\bar{h}_1,0,0,\bar{h}_4,\bar{h}_5)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ inside } f_j\\
(\bar{h}_1,\bar{h}_2,\bar{h}_3,\bar{h}_4,\bar{h}_5)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ outside } f_j\end{array}\right.
\end{equation}
It is straightforward that the above function vanishes if and only if two convex polygons are in contact.

As for grasps, we need to define a parameterized complement constraint for the contact constraint in order to specify the submanifold of configurations that are reachable from one configuration while keeping the object in a constant stable pose. The naive way consists in defining
$$
h_{comp}(\conf) = \left\{\begin{array}{lll}(\bar{h}_2,\bar{h}_3,\bar{h}_6)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ inside } f_j\\
(0,0,\bar{h}_6)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ outside } f_j\end{array}\right.
$$
$\bar{h}_2$, $\bar{h}_3$, $\bar{h}_6$ respectively represent the translation in $x-y$ plane and the rotation around $z$-axis of frame $o_i$ with respect to frame $f_j$. Let $\conf_0$ be a configuration such that $h(\conf_0)=0$. The submanifold defined by
\begin{equation}\label{eq:stable-pose}
\left\{\conf\in\CS,\ \ h(\conf)=0\ \ h_{comp}(\conf) = h_{comp}(\conf_0) \right\}
\end{equation}
contains one object pose for each pair of polygons $(o_i,f_j)$, and there are
$|I| . |J|$ possible combinations. This contraint is therefore not suitable to
enforce object immobility along a path since the object may jump from one pose to another one.

@@

\subsection{constraint graph}

According to Definition~\ref{def:manipulation-problem}, the set of admissible configurations is the union of sub-manifolds of the configuration space of the system. Each sub-manifold is defined by grasp and/or stable contact constraints. We call each sub-manifold a \textit{state} of the problem. A state is defined by a set of grasps and a set of stable contact poses.

A state can be defined by a subset of active grasps, any object not grasped being in a stable contact pose. Let $\ngrippers$ and $\nhandles$ repsectively denote the number of grippers and handles. A state $\state$ is denoted by a vector of size $\ngrippers$:
\begin{equation}\label{eq:state}
\state = \left(h_1,\cdots,h_{\ngrippers}\right)
\end{equation}
where $h_i\in\{\emptyset,1,\cdots,\nhandles\}$ denotes the id of the handle grasped by gripper $i$; $h_i=\emptyset$ means that gripper $i$ does not grasp any handle.

\paragraph{Number of states:} note that for $i\in\{1,\cdots,\nhandles\}$ the number of occurences of $i$ in $\state$ is at most 1: a handle cannot be grasped by several grippers. Note also that the number of occurences of $\emptyset$ is not limited: several grippers may hold nothing. Let $m$ be a non-negative integer not greater than $\ngrippers$ and than $\nhandles$ and let us count the number of states with $m$ handles grasped. The number of subset of $m$ handles among $\nhandles$ is equal to $\frac{\nhandles!}{(\nhandles-m)!m!}$. And the number of ways of dispatching them among the $\ngrippers$ grippers is equal to $\frac{\ngrippers!}{(\ngrippers-m)!}$. Thus, the total number of states is equal to
$$
\sum_{m=0}^{\min(\ngrippers,\nhandles)} \frac{\nhandles!}{(\nhandles-m)!m!} \frac{\ngrippers!}{(\ngrippers-m)!}
$$
\begin{definition}\label{def:neighboring-states}\textbf{Neighboring states}\\
  Two states $\state_1=(h_{1 1},\cdots,h_{\ngrippers 1})$ and $\state_2=(h_{1 2},\cdots,h_{\ngrippers 2})$ are neighboring if they differ by only one grasp and the grasp is empty in one of the states:
  \begin{align*}
    &\exists i\in\{1,\cdots,\ngrippers\},\,h_{i 1} \not= h_{i 2} \mbox{ and } (h_{i 1}=\emptyset \mbox{ or } h_{i 2}=\emptyset), \mbox{ and }\\
    &\forall j\in\{1,\cdots,\ngrippers\},\, j\not=i,h_{j 1} = h_{j 2}.
  \end{align*}
\end{definition}
\begin{definition}\label{def:constraint-graph}\textbf{Constraint graph}\\
  The constraint graph related to a manipulation problem as defined in Definition~\ref{def:manipulation-problem} is a graph
  \begin{itemize}
  \item the nodes of which are states defined by subsets of grasps~(\ref{eq:state}),
  \item an edge connects two states if and only if they are neighboring or they are equal. Edges are also called \textit{transitions}.
  \end{itemize}
  Nodes contain
  \begin{itemize}
  \item the grasp constraints that are active in the corresponding state,
  \item a placement constraint for each object that is not grasped by any handle.
  \end{itemize}
  Transitions contain
  \begin{itemize}
  \item the constraints of the node they connect with the least active grasps,
  \item the parameterized complement constraint of each of the latter.
  \end{itemize}
\end{definition}

\subsection{Example}

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{figures/example.png}
    \def\svgwidth{\columnwidth}
    \graphicspath{{./figures/}}
    \input{figures/example-graph.pdf_tex}
  \end{center}
  \caption{Top: two robot UR3 with one gripper each (X=red, Y=gree, Z=blue) manipulating a cylinder with two handles. The environment contains one rectangular contact surface (in red). The cylinder has two rectangular contact surfaces (in green).
  Bottom: the corresponding constraint graph. Names of states follow Expression~(\ref{eq:state}): for instance, $(\emptyset,1)$ means that gripper of robot 2 grasps handle 1 of the cylinder. In this state, there is no placement constraint.}
  \label{fig:2ur3-cylinder}
\end{figure}
To illustrate the notions exposed in the previous sections, let us consider an example of two UR3 robots manipulating a cylinder illustrated in Figure~\ref{fig:2ur3-cylinder}. The robot is equipped with one gripper attached to the end-effector. The cylinder is equipped with two handles and with two square contact surfaces corresponding to the top and bottom faces of the cylinder.

Grasp constraints are denoted by
$$
grasp_{ij} \mbox{ where } i,j\in\{1,2\}.
$$
Grasp complement constraints are denoted by
$$
grasp_{ij}/comp \mbox{ where } i,j\in\{1,2\}.
$$

$i$ is the index of the gripper $j$ is the index of the handle. The flag of the handles is
$$(\true,\true,\true,\false,\true,\true).$$
Therefore grasp constraints are of dimension 5 and keep the rotation of the gripper around the cylinder axis free.
The object placement constraint is denoted by $place$ and the complement by $place/comp$.
\begin{table}
  \begin{center}
    \begin{tabular}{|p{.44\linewidth}|p{.44\linewidth}|}
      \hline
      state & active constraints\\
      \hline
      $(\emptyset,\emptyset)$ & $place$ \\
      $(j,\emptyset),\,j\in\{1,2\}$ & $grasp_{1j}$\\
      $(\emptyset,j),\,j\in\{1,2\}$ & $grasp_{2j}$\\
      $(i,j)\,i,j\in\{1,2\}$ & $grasp_{1i},grasp_{2j}$\\
      \hline
    \end{tabular}
  \end{center}
  \caption{State constraints.}
  \label{tab:state-constraints}
\end{table}
Table~\ref{tab:state-constraints} indicates which constraints are active for each state.

\begin{table}
  \begin{center}
    \begin{tabular}{|p{.23\linewidth}|p{.18\linewidth}|p{.43\linewidth}|}
      \hline
      transition & belongs to & additional constraints\\
      \hline
      $(\emptyset,\emptyset)\rightarrow (i,\emptyset)$ & $(\emptyset,\emptyset)$ & $place/comp$\\
      $(\emptyset,\emptyset)\rightarrow (\emptyset,i)$ & $(\emptyset,\emptyset)$ & $place/comp$\\
      $(i,\emptyset)\rightarrow(i,j)$ & $(i,\emptyset)$ & $grasp_{1i}/comp$\\
      $(\emptyset,j)\rightarrow(i,j)$ & $(\emptyset,j)$ & $grasp_{2j}/comp$\\
      \hline
    \end{tabular}
  \end{center}
  \caption{Transition constraints: $i,j$ are either 1 or 2. column "belongs to" means that paths along the transition belong to the state.}
  \label{tab:transition-constraints}
\end{table}

@@

  \algblock[globalVariable]{GlobalVariable}{EndGlobalVariable}

@@

    \Comment{gripper indices}

@@

    \Comment{handle indices}

@@

@@

    \State{\Call{recurse}{$\mathcal{G}^{\prime}$,$\mathcal{H}^{\prime}$,$Gr^{\prime}$}}

@@

    \State{$\state.Pl\gets\left[\true,...,\true\right]$}\Comment{list of size $\nobjects$}

@@

@@

@@

    \Function{existState}{$Gr$}
    \Comment{Return $\true$ if a state has already been created for the set of
    grasps given as input.}
    \Function{state}{$Gr$}
    \Comment{Return the state created with the set of grasps given as input.}
    \Function{objectIndex}{$h$}
    \Comment{Return the index of the object handle $h$ belongs to.}

@@

    construction starts by the state with no grasp. Call to recurse
    function loops over the available grippers and handles and create
    states with one more grasp, and a transition to these new
    states. In each state, a placement constraint is added for each
    object of which no handle is grasped. Variables $\mathcal{G}$ and
    $\mathcal{H}$ contain the indices of the free grippers and
    handles. Variable $Gr$ stores the current set of grasps following
    Expression~(\ref{eq:state}). Functions {\scriptsize GRASP}{\small C}{\scriptsize ONSTRAINT}, {\scriptsize GRASP}{\small C}{\scriptsize ONSTRAINT}{\small C}{\scriptsize OMP},{\scriptsize PLACE}{\small C}{\scriptsize ONSTRAINT}, {\scriptsize PLACE}{\small C}{\scriptsize ONSTRAINT}{\small C}{\scriptsize OMP} build grasp, placement constraints and their complements as defined in sections~\ref{subsec:grasp} and ~\ref{subsec:placement}.}

@@

\subsection{M-RRT}

>>>>

\section{Introduction}
This \paper describes a software platform taylored for manipulation planning in
robotics.
The main contributions are:
\begin{itemize}
\item an original and general implementation of prehensile manipulation based on
  non-linear constraints,
\item an original solver for non-linear constraints that is able to handle implicit and explicit constraints,
\item a manipulation planning algorithm that is able to tackle a great variety of manipulation planning problems,
\item an open-source software suite that implements all the above following state of the art development tools and methods.
\end{itemize}

The \paper is organized as follows.
\Partie~\ref{partie:preliminary} introduces some preliminary notions like kinematic chains and Lie groups that are used to model the configuration space of each joint. \Partie~\ref{sec:constraints} introduces non linear constraints and solvers that are at the core of manipulation problem definition. \Partie~\ref{sec:manipulation-problem} defines the problem of prehensile manipulation in the general setting. \Partie~\ref{sec:manipulation-planning} provides a general algorithm that solves manipulation planning problems. Finally, \partie~\ref{sec:humanoid-path-planner} is devoted to a software platform implementing the notions introduced in the previous sections.

Each \partie is implemented by one or several software packages. Instead of providing formulas for some values that need to be computed, we provide a link to the C++ implementation.

\section{Related work}\label{sec:related-work}

Motion planning has given rise to a lot of research work for the past decades.
The problem consists in finding a collision-free path for a given system in
an environment populated with obstacles. The field covers a large variety of
different applications ranging from navigation for autonomous vehicles in
partially known environments~\cite{DarpaUrbanChallenge} to path planning for
deformable objects~\cite{LamKav2001},\\
\cite{RouFerTai2020} and many other applications like coverage path planning~\cite{coverage2001},\\ \cite{coverage2013}, or pursuit evasion planning~\cite{pursuit-evasion1999}.

Planning motions for high dimensional robots like humanoid robots or multi-arm
system has been proven to be of a very high complexity~\cite{SchSha1983},\\ \cite{Canny1983}. Starting in the 1990's random sampling methods have been proposed
to solve the problem, trading the completeness property against efficiency in
solving problem in high dimensional configuration spaces\\ \cite{KSLO1996}, \cite{HsuLatMot99}, \cite{KufLav00}. Since then, asymptotically optimal random sampling algorithms have
been proposed~\cite{KarFra2011}.

\subsection{Path planning with non-linear constraints}

Some systems are subject to non-linear constraints. These non-linear constraints
define a sub-manifold of the configuration the robot must stay on. For instance,
legged robots that must keep contact with the ground and enforce quasi-static
equilibrium, or multi-arm systems grasping a same object are subject to this
type of constraints. As the volume of the constrained manifold is usually equal
to zero, sampling random configurations satisfying the constraints is an event
of zero probability. To sample configurations on the constrained manifold,
\cite{DalNakLamLau2009} and \cite{berenson2011} projects random configurations using a generalization of
Newton-Raphson algorithm. Another way to sample configurations on a constrained
manifold consist in expressing some configuration variables with respect to
others~\cite{Cortes2002},\\ \cite{MirLam2018} when this is possible.\\ \cite{atlasRRT} propose another method based on non-linear projection. They cover the
constrained manifold by growing an atlas composed of local charts. This
approximation provides a probability distribution that is closer to the uniform
distribution over the manifold than the projection of a uniform distribution
over the configuration space.

\subsection{Manipulation planning}

Manipulation planning is a particular instance of path planning were
some objects are moved by some robots. The configuration space of the
whole system is constrained by non-linear constrained due to the fact
that objects cannot move by themselves and should stay in a stable
pose when not grasped by a robot.  The accessible configuration space
is thus a union of sub-manifold as defined in the previous
section. Each of these manifold is moreover a foliation where each
leaf corresponds to a stable position of an object or to a grasp of an
object by a gripper. The geometrical structure of the problem is now
well understood~\cite{AlaSimLau89}. Although several instances of
manipulation problem exist like manipulation by
pushing~\cite{BenRiv1998}, or by throwing~\cite{WooZacLyn2017}, as well as
multicontact planning \cite{Bretl2006}, \cite{LenVaiYosKhe2013},
\cite{TonDelPetParManMan2018}, in this paper, we are only concerned with
prehensile manipulation planning.

Papers about manipulation planning are commonly divided into several categories.

Navigation Among Movable Obstacles (NAMO) \cite{wilfong1988motion}, \cite{stilman2008planning} consists in finding
a path for a robot that needs to displace objects in order to move to a goal
configuration. The final position of the objects does not matter in this case.

Rearrangement planning \cite{ota2004rearrangement}, \cite{LIS210}, consists in finding a sequence of manipulation paths that
move some objects from an initial pose to a final pose. The final configuration
of the robot is not specified. A simplifying asumption is the existence of a
monotone solution, that is a solution where each object is grasped once only and
is moved from its initial pose to its final pose \cite{stilman2008planning}, \cite{stilman2007manipulation}, \cite{srivastavaetal2014icra}, \cite{nieuwenhuisen2008effective}, \cite{ota2004rearrangement}. They mostly use two-level methods
composed of a symbolic or task planner and of a motion planner~\cite{lavalle2006planning}.

\subsection{Open-source software platforms}

@@

\paragraph{Vector representations}
Each Lie group element is represented by a vector. Rotations are represented by
unit quaternions.

@@

\section{Non-linear constraints and solvers}\label{sec:constraints}

@@

To optimize constraint resolution, we perform variable substitution when possible in order to reduce both the number of variables and the dimension of the resulting implicit constraint. We describe here the method that has been first published in~\cite{MirLam2018}. Unlike in the former paper, the description we give in Algorithm~\ref{alg:ExplicitConstraintSet} is closer to the real implementation. Some links to the source code are indeed provided in the algorithm description.
\begin{algorithm}
  \begin{algorithmic}[1]
    \Procedure{initializeSolver}{}
    \State{$explicit$} $\gets$ empty vector of explicit constraints
    \State{$nc \gets 0$}
    \State{$args \gets$ array of size $nq$ filled with -1}
    \EndProcedure
    \Function{add}{$E=(in, out, f)$}
    \If{$args_{out}$ contains an element $\geq 0$}
    \State{return failure}
    \EndIf
    \State{queue $idxArg \gets$ elements of $in$}
    \While{$idxArg$ \textbf{not} empty}
    \State{$iArg \gets idxArg$ first element}
    \State{remove $idxArgs$ first element}
    \If{$iArg\in out$}
    \State{return failure}
    \EndIf
    \If{$args[iArg] == -1$}
    \State{continue}
    \Else
    \State{push $explicit[args[iArg]].in$ elements into $idxArg$}
    \EndIf
    \EndWhile
    \State{fill $args_{out}$ with $nc$}
    \State{explicit.add($E$); $nc\gets nc+1$}
    \State{computeOrder()}
    \State{\Return{success}}
    \EndFunction
  \end{algorithmic}
  \caption{Insertion of an explicit constraint in the solver. Line 1 is called once at initialization of the solver. $explicit$ is a vector that stores the constraints that are successfully added to the solver. $nc$ is the size of the latter. $args$ is an array that stores for each configuration variable, the index in $explicit$ of the constraint that computes this configuration variable, -1 if no constraints computes the index. Procedure \href{https://github.com/humanoid-path-planner/hpp-constraints/blob/2555dfec945575f824bd973c30c5c13ec3b67645/src/explicit-constraint-set.cc\#L144}{ADD} tests whether explicit constraint $E$ is compatible with the previously inserted constraints. Line 6 checks whether any output variable of $E$ is already computed by a previous explicit constraint. If so the procedure returns failure and the $E$ is not inserted. The Loop at line 9 recursively checks that any element of $out$ is not an input variable of a previously inserted constraint. If the loop ends without returning failure,  line 18 stores that elements of $out$ are computed by $E$ and $E$ is inserted in vector of constraints. Function \href{https://github.com/humanoid-path-planner/hpp-constraints/blob/2555dfec945575f824bd973c30c5c13ec3b67645/src/explicit-constraint-set.cc\#L301}{computeOrder} at line 20 recursively computes the order in which the explicit constraints are evaluated, following the rule that the input of a constraint should be evaluated before the output.}
  \label{alg:ExplicitConstraintSet}
\end{algorithm}
Once several compatible explicit constraints have been inserted in the solver,
they behave as a single one. For instance, if $\conf=(\conf_1,\conf_2,\conf_3)$,
$$
  \left\{\begin{array}{lcl}\conf_1 &=& f_1(\conf_2)\\
  \conf_2 &=& f_2(\conf_3)\end{array}\right. \mbox{ becomes }
  \left[\begin{array}{cc}\conf_1\\ \conf_2\end{array}\right] =
    \left[\begin{array}{cc}f_1(f_2(\conf_3))\\ f_2(\conf_3)\end{array}\right],
$$
and $f_2$ should be evaluated before $f_1$.

@@

\section{Manipulation Problem}\label{sec:manipulation-problem}

@@

We denote by $o_i(\conf)$ and $f_j(\conf)$ the poses in configuration $\conf$ of frames with respective centers $C_{o_i}$ and $C_{f_j}$ and with $x$-axis normal to each polygon. In an analogous way as in Definition~\ref{def:grasp}, we define
\begin{equation}\label{eq:convex-shape-contact-hbar}
\bar{h}(\conf) = \log_{\reals^3\times SO(3)} \left(f_{j}(\conf)^{-1}o_i(\conf)\right)
\end{equation}
The contact constraint is defined by the following piecewise differentiable function:
\begin{equation}\label{eq:convex-shape-contact}
h(\conf) = \left\{\begin{array}{lll}(\bar{h}_1,0,0,\bar{h}_4,\bar{h}_5)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ inside } f_j\\
(\bar{h}_1,\bar{h}_2,\bar{h}_3,\bar{h}_4,\bar{h}_5)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ outside } f_j\end{array}\right.
\end{equation}
It is straightforward that the above function vanishes if and only if two convex polygons $o_i$ and $f_j$ are in contact and $C_{o_i}$ is inside $f_j$.

As for grasps, we need to define a parameterized complement constraint for the contact constraint in order to specify the submanifold of configurations that are reachable from one configuration while keeping the object in a constant stable pose. The naive way consists in defining
$$
h_{comp}(\conf) = \left\{\begin{array}{lll}(\bar{h}_2,\bar{h}_3,\bar{h}_6)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ inside } f_j\\
(0,0,\bar{h}_6)&
\mbox{if} & P(C_{o_i},f_j)\mbox{ outside } f_j\end{array}\right.
$$
$\bar{h}_2$, $\bar{h}_3$, $\bar{h}_6$ respectively represent the translation in $x-y$ plane and the rotation around $z$-axis of frame $o_i$ with respect to frame $f_j$. Let $\conf_0$ be a configuration such that $h(\conf_0)=0$. The submanifold defined by
\begin{equation}\label{eq:stable-pose}
\left\{\conf\in\CS,\ \ h(\conf)=0\ \ h_{comp}(\conf) = h_{comp}(\conf_0) \right\}
\end{equation}
contains one object pose for each pair of polygons $(o_i,f_j)$, and there are
$|I| . |J|$ possible combinations. This constraint is therefore not suitable to
enforce object immobility along a path since the object may jump from one pose to another one.

@@

\subsection{Merging constraint and complement into an explicit constraint}
\label{subsec:explicit}

Note that when a grasp constraint and its complement are combined, they
constitute an explicit constraint since the pose of the object grasped
uniquely depends on the configuration of the robot that grasps.

Similarly, when a placement constraint and its complement are combined, they
constitute an explicit constraint since the pose of the object placed uniquely
depends on the pose of the contact surface on which the object is placed. This
latter pose
\begin{itemize}
\item either depends on the configuration of the robot the contact surface
  belongs to,
\item or is constant if the contact surface belongs to the environment.
\end{itemize}
In any case, the explicit expression of the object pose depends on the right
hand side of the complement constraint that is constant along transition paths.

During the construction of the constraint graph (described in Section~\ref{subsec:constraint graph}), grasp and placement
constraints, their complements and the associated explicit constraints
are created together and registered using method\\
\texttt{registerConstraint} of Class
\href{https://gepettoweb.laas.fr/hpp/hpp-manipulation/doxygen-html/classhpp_1_1manipulation_1_1graph_1_1Graph.html}{\texttt{ConstraintGraph}}.

\subsection{Constraint Graph}\label{subsec:constraint graph}

According to Definition~\ref{def:manipulation-problem}, the set of admissible configurations is the union of sub-manifolds of the configuration space of the system. Each sub-manifold is defined by grasp and/or stable contact constraints. We call each sub-manifold a \textit{state} of the problem.

A state can be defined by a subset of active grasps, any object not grasped being in a stable contact pose. Let $\ngrippers$, $\nhandles$, and $\nobjects$ respectively denote the number of grippers, handles, and objects.

We denote by
\begin{itemize}
\item $grasp_{ij}\ i\in\{1,\cdots,\ngrippers\}\ j\in\{1,\cdots,\nhandles\}$ the grasp constraint of handle $j$ by gripper $i$,
\item $grasp_{ij}/comp\ i\in\{1,\cdots,\ngrippers\}\ j\in\{1,\cdots,\nhandles\}$ the complement constraint of the latter,
\item $place_{i}\ i\in\{1,\cdots\nobjects\}$ the placement constraint of object $i$,
\item $place_{i}/comp\ i\in\{1,\cdots\nobjects\}$ the complement constraint of the latter.
\end{itemize}

A state $\state$ is denoted by a vector of size $\ngrippers$:
\begin{equation}\label{eq:state}
\state = \left(h_1,\cdots,h_{\ngrippers}\right)
\end{equation}
where $h_i\in\{\emptyset,1,\cdots,\nhandles\}$ denotes the id of the handle grasped by gripper $i$; $h_i=\emptyset$ means that gripper $i$ does not grasp any handle.

\paragraph{Number of states:} note that for $i\in\{1,\cdots,\nhandles\}$ the number of occurences of $i$ in $\state$ is at most 1: a handle cannot be grasped by several grippers. Note also that the number of occurences of $\emptyset$ is not limited: several grippers may hold nothing. Let $m$ be a non-negative integer not greater than $\ngrippers$ nor $\nhandles$ and let us count the number of states with $m$ handles grasped. The number of subset of $m$ handles among $\nhandles$ is equal to $\frac{\nhandles!}{(\nhandles-m)!m!}$. And the number of ways of dispatching them among the $\ngrippers$ grippers is equal to $\frac{\ngrippers!}{(\ngrippers-m)!}$. Thus, the total number of states is equal to
$$
\sum_{m=0}^{\min(\ngrippers,\nhandles)} \frac{\nhandles!}{(\nhandles-m)!m!} \frac{\ngrippers!}{(\ngrippers-m)!}
$$
\begin{definition}\label{def:neighboring-states}\textbf{Neighboring states}\\
  Two states $\state_1=(h_{1 1},\cdots,h_{\ngrippers 1})$ and $\state_2=(h_{1 2},\cdots,h_{\ngrippers 2})$ are neighboring if they differ by only one grasp and the grasp is empty in one of the states:
  \begin{align*}
    &\exists i\in\{1,\cdots,\ngrippers\},\,h_{i 1} \not= h_{i 2} \mbox{ and } (h_{i 1}=\emptyset \mbox{ or } h_{i 2}=\emptyset), \mbox{ and }\\
    &\forall j\in\{1,\cdots,\ngrippers\},\, j\not=i,h_{j 1} = h_{j 2}.
  \end{align*}
\end{definition}
\begin{definition}\label{def:constraint-graph}\textbf{Constraint graph}\\
  The constraint graph related to a manipulation problem as defined in Definition~\ref{def:manipulation-problem} is a graph
  \begin{itemize}
  \item the nodes of which are states defined by subsets of grasps~(\ref{eq:state}),
  \item an edge connects two states if and only if they are neighboring or they are equal. Edges are also called \textit{transitions}.
  \end{itemize}
  Nodes contain
  \begin{itemize}
  \item the grasp constraints that are active in the corresponding state,
  \item a placement constraint for each object that is not grasped by any handle.
  \end{itemize}
  Transitions contain
  \begin{itemize}
  \item the constraints of the node they connect with the least active grasps,
  \item the parameterized complement constraint of each of the latter.
  \end{itemize}
\end{definition}

\subsection{Example}\label{subsec:ur3-cylinder}

\begin{figure}
  \begin{center}
    \includegraphics[width=\linewidth]{figures/example.png}
    \def\svgwidth{\columnwidth}
    \graphicspath{{./figures/}}
    \input{figures/example-graph.pdf_tex}
  \end{center}
  \caption{Top: two robot UR3 with one gripper each (X=red, Y=gree, Z=blue) manipulating a cylinder with two handles. The environment contains one rectangular contact surface (in red). The cylinder has two rectangular contact surfaces (in green).
  Bottom: the corresponding constraint graph. Names of states follow Expression~(\ref{eq:state}): for instance, $(\emptyset,1)$ means that gripper of robot 2 grasps handle 1 of the cylinder. In this state, there is no placement constraint.}
  \label{fig:2ur3-cylinder}
\end{figure}
To illustrate the notions exposed in the previous sections, let us consider an example of two UR3 robots manipulating a cylinder illustrated in Figure~\ref{fig:2ur3-cylinder}. The robot is equipped with one gripper attached to the end-effector. The cylinder is equipped with two handles and with two square contact surfaces corresponding to the top and bottom faces of the cylinder. $\ngrippers=2,\nhandles=2,\nobjects=1$. The flag of the handles are
$$(\true,\true,\true,\false,\true,\true).$$
Therefore grasp constraints are of dimension 5 and keep the rotation of the gripper around the cylinder axis free.
\begin{table}
  \begin{center}
    \begin{tabular}{|p{.44\linewidth}|p{.44\linewidth}|}
      \hline
      state & active constraints\\
      \hline
      $(\emptyset,\emptyset)$ & $place_1$ \\
      $(j,\emptyset),\,j\in\{1,2\}$ & $grasp_{1j}$\\
      $(\emptyset,j),\,j\in\{1,2\}$ & $grasp_{2j}$\\
      $(i,j)\,i,j\in\{1,2\}$ & $grasp_{1i},grasp_{2j}$\\
      \hline
    \end{tabular}
  \end{center}
  \caption{State constraints.}
  \label{tab:state-constraints}
\end{table}
Table~\ref{tab:state-constraints} indicates which constraints are active for each state. Table~\ref{tab:transition-constraints} indicates which constraints are active for each transition.

\begin{table}
  \begin{center}
    \begin{tabular}{|p{.23\linewidth}|p{.18\linewidth}|p{.43\linewidth}|}
      \hline
      transition & belongs to & additional constraints\\
      \hline
      $(\emptyset,\emptyset)\rightarrow (i,\emptyset)$ & $(\emptyset,\emptyset)$ & $place_1/comp$\\
      $(\emptyset,\emptyset)\rightarrow (\emptyset,i)$ & $(\emptyset,\emptyset)$ & $place_1/comp$\\
      $(i,\emptyset)\rightarrow(i,j)$ & $(i,\emptyset)$ & $grasp_{1i}/comp$\\
      $(\emptyset,j)\rightarrow(i,j)$ & $(\emptyset,j)$ & $grasp_{2j}/comp$\\
      \hline
    \end{tabular}
  \end{center}
  \caption{Transition constraints: $i,j$ are either 1 or 2. column
    "belongs to" means that paths along the transition belong to the
    state, \textit{i.e.} the transition contains the state
    constraints.}
  \label{tab:transition-constraints}
\end{table}

@@

  \algblockdefx[globalVariable]{GlobalVariable}{EndGlobalVariable}{\textbf{global variables}}{}

@@

    \Comment{list of gripper indices}

@@

    \Comment{list of handle indices}

@@

    \State{$isNewState\gets\mathbf{not}\ $\Call{existState}{$Gr^{\prime}$}}

@@

    \If{$isNewState$}{\Call{recurse}{$\mathcal{G}^{\prime}$,$\mathcal{H}^{\prime}$,$Gr^{\prime}$}}\label{cond:recurse}

@@

    \State{$\state.Pl\gets\left[\true,...,\true\right]$}\Comment{list of size $\nobjects$}\label{beg:comp-placement-constraints}

@@

    \EndFor\label{end:comp-placement-constraints}

@@

    \label{beg:set-placement-constraints}

@@

    \EndFor\label{end:set-placement-constraints}

@@

    construction starts by the state with no grasp. Call to
    {\scriptsize RECURSE} function loops over the available grippers
    and handles and creates states with one more grasp, and a
    transition to these new states. In each state, a placement
    constraint is added for each object of which no handle is
    grasped. Variables $\mathcal{G}$ and $\mathcal{H}$ contain the
    indices of the free grippers and handles. Variable $Gr$ stores the
    current set of grasps following Expression~(\ref{eq:state}).
    Lines~\ref{beg:comp-placement-constraints} to
    \ref{end:comp-placement-constraints}
    compute which objects are not grasped.
    Lines~\ref{beg:set-placement-constraints} to
    \ref{end:set-placement-constraints} insert placement constraints in the
    state for those objects. Line~\ref{cond:recurse} recurses only if the
    lastest node reached is new.}

@@

\subsection{Manipulation-RRT}
